From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-30 upi.tex
work on intro to section 3

We now switch gear and focus on constructions of \emph{univalent
  universes}. In order to state the definition of such universes we
need to introduce two notions: propositional equality
$\AgdaDatatype{≡}$ and equivalence $\AgdaDatatype{≃}$.

Propositional equality, or identity types in intensional dependent
type theory, or path types under the HoTT-interpretation come equipped
with a single constructor \AgdaInductiveConstructor{refl} and an
induction principle $\AgdaFunction{J}$. A peculiar aspect of
$\AgdaFunction{J}$ is that it does \emph{not} enforce that every
element of an identity type
$\AgdaFunction{x}~\AgdaDatatype{≡}~\AgdaFunction{y}$ is
\AgdaInductiveConstructor{refl}. This ``weakness'' has a beautiful
consequence: instead of identifying all proofs and collapsing the
structure of proofs in the process, it instead permits a rich
combinatorial structure of proofs to emerge. The second notion we need
is that of equivalence. We say that two types or spaces $A$ and $B$
are equivalent if there exist functions $f : A \rightarrow B$ and
$g : B \rightarrow A$ that compose to the identity function. A notable
point is that equivalences do not come with an induction principle. We
say a universe $\mathcal{U}$ is univalent if its space of identities
$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of
equivalences $\AgdaFunction{A} ≃ \AgdaFunction{B}$.

Instead of talking about univalent universes, we generalize the
discussion to univalent type families which will allow the
constructions of custom universes.

@@

We will consider a general statement of \emph{univalence} for type families. We
start with a some preliminaries. Fig.~\ref{fig:fib} depicts a type family
$P : A \rightarrow \mathcal{U}$ mapping each point $x : A$ to a space $P(x)$. As
everything is continuous, we expect this type family to respect identities. In
other words if $x:A$ and $y:A$ are identified using a path $x \equiv y$, it must
be the case that the spaces $P(x)$ and $P(y)$ become equivalent
$P(x) \simeq P(y)$ as a consequence. Generally speaking, there is no reason to
expect that the spaces $x \equiv y$ and $P(x) \simeq P(y)$ to be themselves
equivalent for all $x$ and $y$. And indeed they are most frequently not
equivalent. In the rare cases in which they are equivalent, the fibration is a
\emph{univalent fibration}.

>>>>

We now switch gear and focus on the construction of \emph{univalent
  universes}. In order to state the definition of such universes we
need to introduce two notions: propositional equality
$\AgdaDatatype{≡}$ and equivalence $\AgdaDatatype{≃}$.

Propositional equality, or identity types in intensional dependent
type theory, or path types under the HoTT-interpretation, come equipped
with a single constructor \AgdaInductiveConstructor{refl} and an
induction principle $\AgdaFunction{J}$. A peculiar aspect of
$\AgdaFunction{J}$ is that it does \emph{not} enforce that every
element of an identity type
$\AgdaFunction{x}~\AgdaDatatype{≡}~\AgdaFunction{y}$ is
\AgdaInductiveConstructor{refl}. This ``weakness'' has a beautiful
consequence: instead of identifying all proofs and collapsing the
structure of proofs in the process, it instead permits a rich
combinatorial structure to emerge.

The second notion we need
is that of equivalence. We say that two types or spaces $A$ and $B$
are equivalent if there exist functions $f : A \rightarrow B$ and
$g : B \rightarrow A$ that compose to the identity function%
\footnote{Strictly speaking, this is a \emph{quasi-equivalence}, an
equivalence also requires a certain coherence between the compositions.}.
A notable point is that equivalences do not come with an induction principle.
We say a universe $\mathcal{U}$ is univalent if its space of identities
$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of
equivalences $\AgdaFunction{A} ≃ \AgdaFunction{B}$.

\jacques{We should probably be precise with our definition of 
equivalence.}

While univalent universes came first, it was later realised that
the more general univalent type families are more convenient to
work with, and allow the constructions of custom universes. 
We thus need to build up some definitions for what this means.
Note that below we will use ``point'' and ``member of a type'',
``space'' and ``type'', ``path'' and ``equivalence'' as
synonyms.  Furthermore our setting is an intuinistic type
theory (with no classical axioms such a choice or excluded
middle added); this implies that all mappings are
\emph{continuous}.

It is very important to remember that while $A$ and $B$ are spaces,
so are $\AgdaFunction{A} ≡ \AgdaFunction{B}$ and
$\AgdaFunction{A} ≃ \AgdaFunction{B}$.  Thus whatever meaningful
questions we can ask of $A$, we can also ask about
$\AgdaFunction{A} ≡ \AgdaFunction{B}$ and
$\AgdaFunction{A} ≃ \AgdaFunction{B}$.

@@

Fig.~\ref{fig:fib} depicts a type family $P : A \rightarrow \mathcal{U}$
mapping each member $x : A$ to a type $P(x)$. As everything is continuous,
we expect this type family to
respect identities. In other words if $x:A$ and $y:A$ are identified using a
path $x \equiv y$, it must
be the case that the spaces $P(x)$ and $P(y)$ are equivalent
$P(x) \simeq P(y)$. Generally speaking, there is no reason to
expect that the spaces $x \equiv y$ and $P(x) \simeq P(y)$ to be themselves
equivalent for all $x$ and $y$. And indeed they are most frequently not
equivalent. In the rare cases in which they are equivalent, the fibration is a
\emph{univalent fibration}.
